---
title: "Visualization with ggplot2"
author: "Presented by William Pang"
date: 'Last Revised: 2022-08-23'
output: html_document
---

```{r setup, include=FALSE}
# Needed for R Markdown
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

*This workshop mostly follows a wonderful [tutorial](https://cengel.github.io/R-data-viz/) created by [Claudia Engel](https://profiles.stanford.edu/claudia-engel), an academic technology specialist and lecturer in Anthropology at Stanford University. However, I've edited Engel's tutorial to place more emphasis behind the syntax logic of `ggplot`.*

Welcome to the Yale StatLab workshop series, where we will be covering data visualization using a tool called `ggplot`, which is part of the `tidyverse` package. When I first learnt `ggplot` on my own, I was quite confused — I came from coding in MATLAB, which was very literal (for instance, if you wanted to plot a sine wave with the equation `y = sin(t)`, it was simply `plot(t,y)`.  

The point of me saying all of this is that if you want to express more complicated ideas, being literal sometimes does not cut it. If a good novel only included terse very manner-of-fact dialogues, you would immediately drop the book. Akin to English, we need a set of rules, a structure of sorts that helps build up a language. This set of "rules" is called grammar, and as you will notice with `ggplot`, there are a lot of idiosyncrasies that might require some blind faith in the beginning, but upon mastery you will begin to see the power of `ggplot`. 

Today, we'll be working with a dataset from the [Stanford Open Policing Project](https://openpolicing.stanford.edu), a project that gathers data across the county on traffic stops made by police officers. In particular, we'll be looking at police stops in Missippi over a span of three years. 

We will use the `read_csv` [function](https://pandas.pydata.org/docs/reference/api/pandas.read_csv.html), which allows for us to define the class types of the data we want read in. 

```{r, results = 'hide'}
# You will need to load up tidyverse. If you do not have the package, use the command install.packages("tidyverse")

library(tidyverse)

# Read the CSV
bias <- read_csv("https://raw.githubusercontent.com/cengel/R-data-viz/master/data/MS_stops_by_county.csv")
```

It's always a good habit to do a quick scan of the data to contextualize what's happening.

```{r, results = 'hide'}
head(bias, n = 2)
```

What's we're looking at is a dataset of policing by counties in Mississippi, with some statistics about the number of Black and White drivers being pulled over. The research question from this dataset is whether there is any racial bias in police stopping -- perhaps we want to plot some sort of scatterplot comparing the percent of Black stopped vs percent of White stopped. 

Before we jump into the coding, the first thing to know about `ggplot` is that it works in layers, sort of like a cake! The bottom layer is your data, and you can layers things on top to make an increasingly sophisticated graph. 

```{r, echo=FALSE}
# Define variable containing url
url <- "https://englelab.gatech.edu/useRguide/images/dataviz/ge_themes.png"
```
<center><img src="`r url`" style = "width: 200px; height: 200px;"></center>

As mentioned, the "base" of your cake is the data.

```{r, eval = FALSE}
ggplot(data = bias)
```

If you just plotted this and pressed run, you will be greeted by a grey screen on the R-studio plot. That's good! We've set the "base" for the cake.

Okay, let's start adding an argument called `aes`. `aes` refers to aesthetic mapping, which is the idea of linking variables in the data to graphical properties in the geometry. If you're working in a cartesian coordinate system, for example, this tells us where the datapoint links exactly on the plane.

Let's add a few more arguments. 

```{r, eval = FALSE}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped))
```

Two things here. First, notice that our arguments for the x and y axis (i.e., `pct_black_stopped` and `pct_white_stopped`) do not have quotation marks on it. This shouldn't come as a surprise to us, especially for those who joined my session last week on Intermediate R, where we introduced the pipe operator. Remember that `ggplot` is under the `tidyverse` universe (no pun intended), similar to how `dplyr` is part of the `tidyverse` universe.

The second question you might ask is "where is my data"? Again, remember we've just *mapped* the data onto the plane, but we haven't told `ggplot` what the graph we want. This is where the fancy stuff begins to come into play. For a scatterplot, where we just want dots or points on the graph, we use the function `geom_point()`.

```{r, results = 'hide'}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped)) +
  geom_point()
```

Notice the `+` sign that we had to use, which means that we're adding an additional layer to our "cake". One thing to note is that the `+` must be added on the same line of your layer -- if you did something like this, for example, it would not work:

```{r, results = 'hide', eval = FALSE}
# This does not work

ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped)) 
  + geom_point()
```

It also doesn't really matter if the `aes` argument goes into the base layer or to the `geom_point` layer.

```{r}
ggplot(data = bias) +
  geom_point(aes(x = pct_black_stopped, y = pct_white_stopped))
```

However, for consistency (and readability), we will stick with putting the `aes` argument in the data layer. 

You can play around with the `geom_point` argument — for instance, you don't want circles on your graph, but want squares instead. And while we're at it, you, the picky customer, might think the squares are too dark, so you would like to change the opacity.

```{r}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped)) +
  geom_point(shape = 'square', alpha = 0.3) # Notice that we need to put quotations on 'square', as the argument 'square' is not piped from our dataset
```

Anyway, let's get back to our original research question, which is to see if there is any bias in police stops. To do this, we should plot a 45$^\circ$ degree reference line (or a slope of m = 1 because y = x) on the scatterplot. If the dots fall on the reference line, this means that there is no bias (e.g., in a county where 30% of Whites are stopped and 30% of Blacks are also stopped, the dot would fall on the reference line). 

We can achieve this with the `geom_abline` function, which draws a line from a to b.

```{r}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = 'red')
```

Hmmm, something seems to be off, as what I wanted was the line to cut diagonally on the plane. This is because the y and x axis are scaled differently - if we wanted the slope to cut diagonally, the two axis have to be the same. We can do this using the `scale_x_continuous` and `scale_y_continous` argument.

```{r}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) 
```

Let's say we were doing some exploratory data analysis and wanted to see data points where less than 10% of White drivers were stopped. Let's trying highlighting these data points by changing the color on our graph. 

```{r}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped, color = pct_white_stopped < 0.1)) + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_point()
```

You might have initially thought that the `color` argument would go into the geometries layer. However, this is not the case, because what we're *conceptually* doing is creating a new column in our dataset, creating a TRUE/FALSE statement (i.e., whether the percent of Whites stopped is greater than 0.1), and then mapping it onto the plane using the `geom_point` function. 

Similarly, if we wanted to change something about the properties of the dots themselves (say we wanted to change the color), we would put our arguments into the `geom_point` function.

```{r}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped)) +
  geom_point(color = 'steelblue')
```

Interesting! Just by visual inspection we see that there's something interesting going on there, as a lot of points seem to be aggregated below the reference line. We can "zoom in" by tweaking `scale_x_continuous` and `scale_y_continuous`:

```{r}
ggplot(data = bias, aes(x = pct_black_stopped, y = pct_white_stopped)) + 
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  scale_x_continuous(limits = c(0, 0.1)) +
  scale_y_continuous(limits = c(0, 0.1)) +
  geom_point()
```

Notice that we also get a warning that some values are cut of because we have essentially cropped our graph.

Really, we've covered the hardest part of the workshop, which is getting used to the language and syntax of `ggplot`. Once we know how the toolkit belt works, what we're really doing now is adding more and more tools into the belt. So give yourselves a pat in the back.

While the scatterplot gives us a good overview, we might want to look at each of the counties in Mississippi. We can do this by finding the difference between `pct_white_stopped` and `pct_black_stopped` (i.e., `pct_white_stopped - pct_black_stopped`), and to see whether there is bias in police stops at each county. This difference has been calculated and stored under the variable `wb_delta`, where a positive $\Delta$ indicates a bias towards stopping Whites.

```{r}
ggplot(data = bias, aes(x = county_name, y = wb_delta)) + 
  geom_col()
```

Clearly, we can't distinguish one county from another county - it's not readable at all. We can fix this problem by flipping the coordinates:

```{r}
ggplot(data = bias, aes(x = county_name, y = wb_delta)) +
  geom_col() + 
  coord_flip()
```

Since most people read from top to bottom, we might want to flip the ordering of `county_name` so that Adams County is on the bottom (or *further top of the screen* — but in graphics the coordinate system is flipped on the y-axis). Unfortunately, there is no intuitive solution (such as adding a negative sign before the `county_name`). You must use the function `scale_x_discrete` and reverse the limits. Note that this will not work if the x variable is continuous. 

```{r}
ggplot(data = bias, aes(x = county_name, y = wb_delta)) +
  geom_col() + 
  coord_flip() +
  scale_x_discrete(limits = rev)
```

Let's try doing the same thing for the continuous variable `wb_delta`, which requires some knowledge of the idiosyncrasies of `ggplot` functions. Again, stackoverflow and Google is your friend here.

```{r}
ggplot(data = bias, aes(x = county_name, y = wb_delta)) +
  geom_col() + 
  coord_flip() +
  scale_y_reverse(limits = c(1,-1)) # You could also just use scale_y_reverse() 
```

We can also reorder the counties by the `wb_delta` values, from lowest (indicating Black bias) to highest (indicating White bias).

```{r}
ggplot(bias, aes(x = reorder(county_name, -wb_delta), y = wb_delta)) +
  geom_col() + 
  coord_flip()
```

Before I forget to mention, we could also simplify the code a bit (which I should have done previously, but for illustrative purposes I copied everything out) by assigning everything before `geom_col()` to `p`.

```{r}
p <- ggplot(bias, aes(x = reorder(county_name, -wb_delta), y = wb_delta)) +
  geom_col() 

p + coord_flip()
```

Okay, so I think we've exhausted this dataset  - let's work with an extension of the police stops dataset which we will call `reasons`.

```{r, results = 'hide'}
reasons <- read_csv("https://github.com/cengel/R-data-viz/raw/master/data/MS_stops.csv")
```

Again, it's a good habit to look at both the `spec` output from `read_csv` and some techniques I covered last week to get a good overview of the data.

```{r, results = 'hide'}
head(reasons, n = 2)
```

Another way to get to know your data is to build a bar graph, especially when we're working with categorical variables. This can be done with the `geom_bar` function.

```{r}
ggplot(data = reasons, aes(x = violation)) + 
  geom_bar()
```

Notice that we don't need to specific a `y` argument for `aes`, which makes sense because we are trying to tabulate the different violation counts. What's happening here is that the `violation` variable gets passed to `geom_bar()`, which basically does the job of `group_by` and `count`. This is something we covered last week:

```{r, eval= FALSE}
reasons %>% group_by(violation) %>% count()
```

Generally, `ggplot` has a shortcut construction that can do what we just illustrated (in the above example, it was `geom_bar`).  But this is not *always* the case, and as such, it's good to know functions like `after_stat`, where you can manually tweak the calculations. For instance, I might be interested in getting the percentage of each of the violations:

```{r}
# Convert to percentage
ggplot(data = reasons, aes(x = violation, y = after_stat(count/ sum(count) * 100))) + 
  geom_bar()
```

We could also change the colors of the bar — but instead of using the `color` function (which just creates a green border for each bar), we can use `fill`. 

```{r}
ggplot(data = reasons, aes(x = violation)) + 
  geom_bar(fill = "green")
```

But just changing the colors is not that insightful. Say, for instance, we wanted to know how many of speeding violations were incurred by Whites compared to Blacks. 

```{r}
ggplot(data = reasons, aes(x = violation)) + 
  geom_bar(aes(fill = driver_race))
```

Here the `aes` function needs to be called again (i.e., `geom_bar(fill = driver_race)` won't work). This is similar to the earlier example, where we wanted to change the `pct_white_stopped` colors after we defined a certain threshold. We could just have included the `fill` argument in our `data` layer. 

```{r, eval = FALSE, fig.hide = TRUE}
ggplot(data = reasons, aes(x = violation, fill = driver_race)) + 
  geom_bar()
```

One more point on this. We could also tell ggplot to stretch the bars between [0,1], and find the proportions of violations that are grouped by race.

```{r}
# Convert to percentage
ggplot(data = reasons, aes(x = violation)) + 
  geom_bar(aes(fill = driver_race), position = "fill")
```

We'll now do a gear shift into plotting time series data, which is dear and near to me as I've worked on analyzing COVID datasets. In this example, we're interested in knowing what the type of violations occur on a given day of the week. 

To do so, we want to split our datasets into days of the week, and then split that dataset based on violations. We should now be familiar with the handy `group_by` function.

```{r}
dow_violations <- reasons %>% 
  group_by(wk_day, violation) %>% 
  count()

head(dow_violations, n=6)
```

Now, let's try plotting it out.

```{r}
ggplot(dow_violations, aes(x = wk_day, y = n)) +
  geom_line()
```

Woah! This wasn't what we wanted! What happened here is that `ggplot` displayed the range of values for each day of the week (on Friday, for example, we see the range to be [564 (Breaks-Lights-etc), 20767 (Speeding)]. Let's fix this by adding the `group` argument.

```{r}
ggplot(dow_violations, aes(x = wk_day, group = violation, y = n)) +
  geom_line()
```

That's better. But we don't know which line corresponds to which violation. To do so, we need to use the `color` function.

```{r}
ggplot(dow_violations, aes(x = wk_day, group = violation, y = n, color = violation)) +
  geom_line()
```

If you notice the days of the week, they are a bit messed up. As we've seen a few times before, R usually arranges categorical variables (even date class) by alphabetical order. To fix this problem, we need to manually specify what the order should be by forcing it as a `factor`. Remember that whenever you need variables to be arranged in a particular order, use factors as factors not only store the variables of interest, but assigns an index to them.

```{r}
dow_violations$wk_day <- factor(dow_violations$wk_day, 
                       levels=c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
```

Let's give this another try.

```{r}
ggplot(dow_violations, aes(x = wk_day, group = violation, y = n, color = violation)) +
  geom_line()
```

This is a great segue to `facet_wrap`, our last subject of the day. Imagine you wanted separate panels for each violation classification. What faceting essentially does is split one big plot into multiple smaller plots, making the plot less confusing and easier to interpret for the viewer.

Let's use the `facet_wrap` function.

```{r}
ggplot(dow_violations, aes(x = wk_day, y = n, group = violation)) +
     geom_line() +
     facet_wrap(~ violation)
```

Notice the `~` tilde. The tilde can be read as "by" - semantically, this would be equivalent to [saying](https://bookdown.org/yih_huynh/Guide-to-R-Book/facet-wrapping.html) "I want to make new graphs separated by the cut categories". 

We can force the facet wrap to have 2 columns, i.e., a 3x2 grid. Let's also rename the `y` label to something else other than n and remove the `x` label because the day of week is pretty intuitive.

```{r}
p <- ggplot(dow_violations, aes(x = wk_day, y = n, group = violation)) +
     geom_line() +
     facet_wrap(~ violation, ncol = 2) +
     labs(title = 'Observed violations per day of week',
     x = "",   # Use "" to remove text on label
     y = "Number of violations")
```

Lastly, you can do a small touch of customization to the facet wrapped graphs using different themes. 

```{r}
p + theme_bw()
```

$\underline{Acknowledgements}$ 

As mentioned in the preamble, I relied mainly on [Claudia Engel's ](https://profiles.stanford.edu/claudia-engel) R tutorial [series](https://cengel.github.io/R-data-viz/). Another useful resource I found was Thomas Lin Pedersen's two-part Youtube [tutorial](https://youtu.be/h29g21z0a68), which helped lay a theoretical foundation on the syntax logic behind ``ggplot``. 